A primer to numerical simulations: The perihelion motion of Mercury

### 2017-05-22 tex/MercuryNumerical.tex
Updated the Numerical implementation section and added an example script (which still needs to be adjusted to the text).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

@@

We can say that we understand a physical system if we can demonstrate that an assumed force leads to the trajectories
observed, where here trajectory means that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

@@

\section{How to treat differential equations on a computer}

@@

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
With this Newtons second law reads 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
\label{eq:newton}
\end{equation}
\begin{itemize}
\item{\bf First task:}
Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
trajectories of Mercury for different start parameters.
Why is it sufficient to work in two dimensions using simply 
$\vec x(t)=(x(t),y(t))^T$?
 When the parameters are chosen appropriately close
orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
of planet and sun) stays fixed in space.
\end{itemize}
It is a well known feature of a $1/r^2$ force that elliptic orbits with fixed perihelion emerge. However, as soon
as the potential is different, the perihelion moves. This will be studied next:
\begin{itemize}
\item{\bf Second task:}
Now the students should add a new term to the right hand side of Eq.~(\ref{eq:newton}). In particular the equation should now
read 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
\label{eq:newton}
\end{equation}
where $\alpha$ is some parameter. What happens to the trajectories as $\alpha$ is varied?
\end{itemize}
Finally the start values need to be chosen appropriately such that the simulation indeed mimicks the trajectories
of Mercury ...

@@

\appendix
\section{Instructions to install V-phython}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}

\usepackage{graphicx}           % Import Pictures + PDFs
\usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
\definecolor{fzjblue}{HTML}{005B81}
\definecolor{light-gray}{gray}{0.97}
\definecolor{mygreen}{rgb}{0,0.6,0}
\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
\lstset{
  language=Python,
  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\linespread{1}\footnotesize,        % the size of the fonts that are used for the code
  commentstyle=\color{mygreen},
  frame=single,
  keywordstyle=\color{blue},       % keyword style
  tabsize=4,	                   % sets default tabsize to 2 spaces
  rulecolor=\color{light-gray},
}

\newcommand{\python}[0]{\texttt{Python} }
\newcommand{\abs}[1]{\left\vert #1 \right\vert}

\begin{document}

@@

We can say that we understand a physical system if we can demonstrate that an assumed force leads to the trajectories
observed, where here trajectory means that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

@@

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
With this Newtons second law reads 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
\label{eq:newton}
\end{equation}
\begin{itemize}
\item{\bf First task:}
Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
trajectories of Mercury for different start parameters.
Why is it sufficient to work in two dimensions using simply 
$\vec x(t)=(x(t),y(t))^T$?
 When the parameters are chosen appropriately close
orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
of planet and sun) stays fixed in space.
\end{itemize}
It is a well known feature of a $1/r^2$ force that elliptic orbits with fixed perihelion emerge. However, as soon
as the potential is different, the perihelion moves. This will be studied next:
\begin{itemize}
\item{\bf Second task:}
Now the students should add a new term to the right hand side of Eq.~(\ref{eq:newton}). In particular the equation should now
read 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter. What happens to the trajectories as $\alpha$ is varied?
\end{itemize}
Finally the start values need to be chosen appropriately such that the simulation indeed mimicks the trajectories
of Mercury ...

\section{Numerical Implementation}
In this section we present the numerical implementation of the perihelion motion of Mercury.  We select \python as the programming language of choice \cite{}, because \python is easy to learn, intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction how to set up python on different operating systems.}.  No prior knowledge of \python or any other programming language is required because we explain all the essential steps in order to create the simulation.  We also provide a working example \cite{} which can be used as template.

The first step of a numerical simulation is the extraction of the relevant physical parameters and quantities in order to translate the problem to mathematical equations.  The relevant objects for this example are the sun and mercury.  Thus, to describe their motion, one needs their "initial" distances, velocities and accelerations which are the input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).  In the case of this problem, one can simplify the implementation by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.  \textit{Why is this simplified picture sufficient to describe the motion}?

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39 km/s $\footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in their respective unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ and distances in $R_0 = 10^{10}m$.  With this choice, the initial distance of mercury to the sun, the initial velocity of mercury and the acceleration prefactor become
\begin{equation}
	r_{MS}(0) = 4.6 R_0 \, , \quad
	v_{M}(0)  = 0.34 \frac{R_0}{T_0} \, ,  \quad
	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
	\, .
\end{equation}

\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
\end{figure}

In python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
Next, we set up the initial vectors which will describe the motion in the two-dimensional plane.  We will build on the existing \python module: \texttt{Vpython}, which provides implementation for the vectorial implementation and animation.  The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions, the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rS0 = vector(0, 0, 0)
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.  In first approximation we can numerically express the updates of the position and the velocity by
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" -- Don't Repeat Yourself)
\begin{lstlisting}
	# Define the function
	def evolve_mercury(vec_rM_old, vec_vM_old):
		<...Code...>
		return vec_rM_new, vec_vM_new

	# Call the function
	vec_rM_new, vec_vM_new = do_time_step(vec_rM_old, vec_vM_old)
\end{lstlisting}
Because of \python{}s syntax, it is necessary that the body of the function is indented relative to the definition statement.

Finally, we can express the evolution by a \texttt{while}-loop
\begin{lstlisting}
	t = 0
	# Execute the loop as long as t < T
	while t < T:
		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM)
		t = t + dt
\end{lstlisting}
Note the required indent of the loop structure similar to the indent of a function.  In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration. The time step \texttt{dt} needs to be sufficiently small in order to avoid numerical errors.  We recommend $dt = ...$.  The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.  To describe at least one full orbit, $T$ needs to be larger than $T>..$.

Ideas for questions:
\begin{itemize}
	\item Mass and motion of sun
	\item Estimate \texttt{dt} and also \texttt{T} in order to cover roughly two "mercury years".
	\item What does happen when you change those units?
	\item Compare values to the values in \cite{}.  Where does the difference come from?
	\item Start with $\alpha=0$. What do you observe?
\end{itemize}

@@

@@

\appendix
\section{Instructions to install V-phython}
\label{appendix:python}
